Describing *Symbiodinium* metacommunities using high-throughput sequencing

### 2017-01-05 manuscript/manuscript.Rmd
prepare draft to share

  The observed structure of *Symbiodinium* communities in a variety of host species was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data. Specifically, whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset or independently within each sample affected the number and taxonomic assignment of OTUs. Clustering sequences from all samples together reduced the total number of OTUs by 98% relative to no clustering (Table 1), while within-sample clustering resolved intermediate diversity, with 13% more OTUs than across-sample clustering. Comparing taxonomic assignments of the OTUs generated by these approaches demonstrates that across-sample clustering in many cases assigns the same dominant OTU to samples with different dominant sequence variants (Fig. 2), while within-sample clustering assigns different OTUs that reflect the dominant variant within that particular sample. This occurs because an OTU's assigned identity is based on its most abundant member sequence, and when members occur across many samples (i.e., during across-sample clustering), each member's abundance reflects the number (and sequencing depth) of samples in the dataset that contain it, with sequences from more heavily-sampled and/or deeply-sequenced hosts being better represented. The best-represented sequences then subsume closely-related but different sequences from other samples into OTUs that bear their names, even if these other samples do not actually contain the named sequences. For ITS2 OTUs, such merging across samples is undesirable since even single nucleotide differences in a sample's dominant ITS2 variant may indicate different *Symbiodinium* species. Therefore, even though clustering at 97% similarity successfully collapses intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when such clustering is applied across a dataset potentially comprised of many closely related types, it is likely to also collapse interspecific variation and underestimate the true diversity of *Symbiodinium* present. Indeed, this limitation was also encountered by @Arif:2014gk when clustering closely-related clade C sequences across samples: despite being comprised mostly of C41 variants, ~3k sequences from *Acropora hemprichii* were subsumed into a single OTU that was named C1 after the dominant variant among the ~7k sequences from *Pocillopora verrucosa* with which they were clustered. These cases clearly demonstrate how clustering sequences across samples causes the identity of OTUs in any given sample to be driven largely by the other samples in the dataset.

  Meanwhile, clustering sequences within each sample independently assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the dataset. The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects knowledge of *Symbiodinium* diversity and ecology that has been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa. Indeed, only the within-sample clustering approach assigned different dominant clade C OTUs to samples of *O. annularis* and *S. siderea*, which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx]. Therefore, when analyzed using within-sample clustering, ITS2 metabarcoding data may generate a good approximation of species-level diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically abundant sequence variants to assign taxonomy. Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa. Therefore, we recommend a within-sample clustering approach for metabarcoding studies where many different *Symbiodinium* types are likely to be encountered.

  Importantly, the assumptions underlying a within-sample clustering approach -- that samples typically contain one *Symbiodinium* clone that can be diagnosed by its most abundant intragenomic ITS2 variant -- will not always be met. Indeed, multiple *Symbiodinium* co-occur frequently in coral samples [@Silverstein:2012ub], which undermines support for assuming that variation within a sample is intragenomic. However, co-occurring types in many cases are members of different clades, whose ITS2 sequences are divergent enough to be resolved separately by 97% clustering. Thus, only when very closely related types co-occur in a sample [e.g., @Sampayo:2007p4020; @Wham:2016gl] is this approach likely to fail. Other problematic cases may occur when multiple intragenomic sequence variants are co-dominant (i.e., comprise nearly equal proportions of the rDNA array), such that slight differences in their relative abundance may lead to different OTU assignments in samples with otherwise highly similar sequence assemblages (e.g., Fig. SX). These challenges may be overcome by developing criteria for taxonomic assignment that are more complex than choosing the single most abundant variant, for example, by taking into account the presence and abundance of multiple variants and/or the phylogenetic relationships among sequences (e.g. to use more evolutionarily derived variants for taxonomic assignment).

  While within-sample clustering of ITS2 data cannot provide species-level descriptions of *Symbiodinium* communities (since the ITS2 marker itself does not provide this resolution), this nuanced bioinformatic approach nevertheless accommodates known evolutionary complexities in the ITS2 locus to provide community descriptions that better approximate true biological diversity than either across-sample clustering or no clustering at all. When treated with an appropriate bioinformatic approach, ITS2 metabarcoding can provide a comprehensive and quantitative picture of *Symbiodinium* taxa present in a sample with a high sensitivity for detection of rare 'background' taxa (which may be ecologically important), and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.

@@

  While environmental data revealed significant biophysical differences between the north and south shores of the island (e.g., wave exposure, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the corals sampled. This may partly be due to low statistical power associated with small sample sizes, since two species showed clear trends: *S. siderea* and *P. furcata* were dominated more frequently by *Symbiodinium* in clades D and A, respectively, on the south shore, compared to the north shore (Fig. 2). Since these symbionts are typically associated with warmer and more variable temperature regimes [e.g., @Oliver:2011cr] and high-light habitats [@Byler], it is consistent with expectation that they were more prevalent on the south shore, where temperatures were slightly higher and chlorophyll levels were lower, suggesting light penetration is likely greater. However, the lack of differentiation in the symbiont communities of most host species suggests their structure is not sensitive to the level of environmental variation between shores.

  When the associations between *Symbiodinium* types and coral species were visualized by network analysis, features of both symbiont and host ecology emerge. First, ITS2 type B1 was "abundant" (i.e. >1%) in individuals of seven out of ten coral species, and "dominant" (i.e. >50%) in five out of ten, suggesting it is a more generalist symbiont lineage. However, analysis of more variable chloroplast and microsatellite loci has revealed that ITS2 type B1 in the Caribbean is comprised of numerous specialized lineages that demonstrate specific associations with different coral host species [@Santos:2004p4044; @Finney:2010p8324; @Parkinson:2015gt]. Likewise, the apparent generalists *Symbiodinium* C3 (which dominated colonies of both *S. siderea* and *M. cavernosa*) and *Symbiodinium* A4 (which dominated both *P. astreoides* and *P. furcata*) may also be comprised of multiple specialized lineages [@Thornhill:2013dx], which will require evaluation with population genetic markers.

@@

  Finally, to reveal symbionts that consistently occupied a 'background' niche, we identified OTUs present at <1% relative abundance in at least three different host species. Such a distribution suggests that these symbionts can occupy a range of hosts, but are unlikely to dominate them under prevailing environmental conditions. Their presence, however, indicates that they may be able to achieve dominance within a colony under certain conditions. Indeed, the most prevalent 'background' symbiont in St. John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io], present in seven out of ten host species), which has been shown to proliferate within hosts during and after thermal stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364; @Silverstein:2015er], or in more marginal environments [@LaJeunesse:2010p7790]. Association with these symbionts may therefore broaden a host's fundamental niche space and increase its resilience to environmental change. *Symbiodinium* A4 and C3 also commonly occurred as background symbionts, suggesting they may perform analogous functional and ecological roles within the metacommunity.

  Focusing on hosts, there was a clear difference among species in the number of OTUs that were abundant (>1% relative abundance, Fig. 4) and dominant (>50% relative abundance, Fig. 5A), indicating different levels of variability in the symbiosis ecology of different coral species. *D. strigosa* individuals were dominated by four different OTUs, while *D. cylindrus*, *S. radians*, *M. cavernosa*, and *P. astreoides* were dominated by a single OTU. In fact, metabarcoding data enables quantifying symbiosis flexibility as multivariate dispersion (a metric of beta diversity), which can be statistically compared among species [@Anderson:2006fs]. The flexibility of symbiosis is important as a potential indicator of the environmental sensitivity of a species [@Putnam:2012wv], and its ability to switch or shuffle its symbionts. This metric revealed that *M. cavernosa*, *P. astreoides*, *S. radians*, and *D. cylindrus* all had the lowest symbiotic flexibility. On the other hand, *D. strigosa*, *P. furcata*, *M. alcicornis*, and *S. siderea* had the most variable symbiont communities, indicating that these species may be more 'flexible' in their associations, and that their symbiont community structure may be more responsive to variability in the natural environment.

@@

  While not providing species-level resolution, ITS2 metabarcoding represents a powerful approach to quantitatively and comprehensively describe *Symbiodinium* metacommunity composition on coral reefs. With proper bioinformatic interpretation, these datasets can address critical topics in coral symbiosis ecology, including the influence of the environment on community structure, changes in dominant or background symbiont abundance over environmental gradients or over time, with important implications for understanding coral responses to environmental change.

>>>>

  The observed structure of *Symbiodinium* communities in a variety of host species was significantly impacted by the bioinformatic approach used to analyze ITS2 sequence data. Specifically, whether sequences were clustered into operation taxonomic units (OTUs) collectively across the entire dataset or independently within each sample affected the number and taxonomic assignment of OTUs. Clustering sequences from all samples together reduced the total number of OTUs by 98% relative to no clustering (Table 1), while within-sample clustering resolved intermediate diversity, with 13% more OTUs than across-sample clustering. Comparing taxonomic assignments of the OTUs generated by these approaches demonstrates that across-sample clustering in many cases assigns the same dominant OTU to samples with different dominant sequence variants (Fig. 2), while within-sample clustering assigns them distinct OTUs that reflect the dominant variant in the sample. This occurs because OTU identity is based on the most abundant member sequence, and when members occur across many samples (i.e., during across-sample clustering), each member's abundance reflects the number (and sequencing depth) of samples in the dataset that contain it, with sequences from more heavily-sampled and/or deeply-sequenced hosts being better represented. The best-represented sequences then subsume closely-related but different sequences from other samples into OTUs that bear their names, even if these other samples do not actually contain the named sequences. For ITS2 OTUs, such merging across samples is undesirable since even single nucleotide differences in a sample's dominant ITS2 variant may indicate different *Symbiodinium* species.

  Therefore, even though clustering at 97% similarity successfully collapses intragenomic variation within a single *Symbiodinium* type [@Arif:2014gk], when such clustering is applied across a dataset potentially comprised of many closely related types, it is likely to also collapse interspecific variation and underestimate the true diversity of *Symbiodinium* present. Indeed, this limitation was also encountered by @Arif:2014gk when clustering closely-related clade C sequences across samples: despite being comprised mostly of C41 variants, ~3k sequences from *Acropora hemprichii* were subsumed into a single OTU that was named C1 after the dominant variant among the ~7k sequences from *Pocillopora verrucosa* with which they were clustered. These cases clearly demonstrate how clustering sequences across samples causes the identity of OTUs in any given sample to be driven largely by the other samples in the dataset.

  Meanwhile, clustering sequences within each sample independently assigns OTUs that reflect only the assemblage of sequences within that sample, and therefore does not depend on the presence or abundance of sequences from other samples in the dataset. The outcome of within-sample clustering of ITS2 sequences as applied herein more accurately reflects patterns of *Symbiodinium* diversity and ecology that have been revealed by more variable markers (psbA^ncr^) and microsatellites [@Finney:2010p8324; @Thornhill:2013dx]; namely, that clades B and C in the Caribbean comprise numerous different *Symbiodinium* species that tend to associate with different host taxa. Indeed, only the within-sample clustering approach assigned different dominant clade C OTUs to samples of *O. annularis* and *S. siderea*, which have been differentiated based on other markers and microsatellites [@Thornhill:2013dx]. Therefore, when analyzed using within-sample clustering, ITS2 metabarcoding data may generate a good approximation of species-level diversity in most cases, and in theory should reflect patterns that would be observed by sequencing dominant ITS2 bands from denaturing gradient gel electrophoresis (DGGE; the most commonly use method of describing *Symbioinium* diversity prior to metabarcoding), since both methods rely on numerically dominant sequence variants to assign taxonomy. Furthermore, metabarcoding overcomes the primary limitations of DGGE by providing quantitative data and highly sensitive detection of low abundance taxa. Therefore, we recommend a within-sample clustering approach for metabarcoding studies where many different *Symbiodinium* types are likely to be encountered.

  Importantly, the assumptions underlying a within-sample clustering approach -- that samples typically contain one *Symbiodinium* clone that can be diagnosed by its most abundant intragenomic ITS2 variant -- will not always be met. Indeed, multiple *Symbiodinium* co-occur frequently in coral samples [@Silverstein:2012ub], which undermines support for assuming that variation within a sample is intragenomic. However, co-occurring types in many cases are members of different clades, whose ITS2 sequences are divergent enough to be resolved separately by 97% clustering. Thus, only when very closely related types co-occur in a sample [e.g., @Sampayo:2007p4020; @Wham:2016gl] is this approach likely to fail. Other problematic cases may occur when multiple intragenomic sequence variants are co-dominant (i.e., comprise nearly equal proportions of the rDNA array), such that slight differences in their relative abundance may lead to different OTU assignments in samples with otherwise highly similar sequence assemblages (e.g., Fig. SX). These challenges may be overcome by basing taxonomic assignment on more complex criteria than just the most abundant variant, such as multiple co-dominant variants, or more evolutionarily derived variants.

  While no treatment of ITS2 data can provide species-level descriptions of *Symbiodinium* communities (since the marker itself does not provide this resolution), within-sample clustering nevertheless accommodates known evolutionary complexities in the ITS2 locus to provide community descriptions that better approximate true biological diversity than either across-sample clustering or no clustering at all. When treated with an appropriate bioinformatic approach, ITS2 metabarcoding can provide a comprehensive and quantitative picture of *Symbiodinium* diversity, and can be easily applied across host species to rapidly survey the *Symbiodinium* metacommunity on a reef-wide scale.

@@

  While environmental data revealed significant biophysical differences between the north and south shores of the island (e.g., wave exposure, temperature, and chlorophyll A), there were no significant differences in symbiont community composition in the host species sampled on either shore. This may partly be due to low statistical power associated with small sample sizes, since two species showed clear trends: *S. siderea* and *P. furcata* were dominated more frequently by *Symbiodinium* in clades D and A, respectively, on the south shore, compared to the north shore (Fig. 2). Since these symbionts are typically associated with warmer and more variable temperature regimes [e.g., @Baker:2004p21; @Oliver:2011cr] and high-light habitats [@LaJeunesse:2002p3917], it is consistent with expectations that they were more prevalent on the south shore, where temperatures were slightly higher, and chlorophyll levels were lower, suggesting greater light penetration. However, the lack of differentiation in the symbiont communities of most host species suggests their structure is not sensitive to environmental differences between shores.

  When the associations between *Symbiodinium* types and coral species were visualized by network analysis, features of both symbiont and host ecology emerge. First, ITS2 type B1 was "abundant" (i.e. >1%) in seven out of ten coral species, and "dominant" (i.e. >50%) in five out of ten, suggesting it is a more generalist symbiont lineage. However, analysis of more variable chloroplast and microsatellite loci has revealed that ITS2 type B1 in the Caribbean is comprised of numerous specialized lineages that demonstrate specific associations with different coral host species [@Santos:2004p4044; @Finney:2010p8324; @Parkinson:2015gt]. Likewise, the apparent generalists *Symbiodinium* C3 (which dominated colonies of both *S. siderea* and *M. cavernosa*) and *Symbiodinium* A4 (which dominated both *P. astreoides* and *P. furcata*) may also be comprised of multiple specialized lineages [@Thornhill:2013dx], which will require evaluation with higher resolution genetic markers.

@@

  Finally, to reveal symbionts that consistently occupied a 'background' niche, we identified OTUs present at <1% relative abundance in at least three different host species. Such a distribution suggests that these symbionts can occupy a range of hosts, but are unlikely to dominate them under prevailing environmental conditions. Their presence, however, indicates that they may be able to achieve dominance under certain conditions. Indeed, the most prevalent 'background' symbiont in St. John was a member of clade D (*Symbiodinium trenchii* [@LaJeunesse:2014io], present in seven out of ten host species), which has been shown to proliferate within hosts during and after thermal stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364; @Silverstein:2015er], or in more marginal environments [@LaJeunesse:2010p7790]. Association with these symbionts may therefore broaden a host's fundamental niche space and increase its resilience to environmental change [@Correa:2010p7787]. *Symbiodinium* A4 and C3 also commonly occurred as background symbionts, suggesting they may perform analogous functional and ecological roles within the metacommunity.

  Focusing on hosts, there was a clear difference among species in the number of OTUs that were abundant (>1% relative abundance, Fig. 4) and dominant (>50% relative abundance, Fig. 5A), indicating different levels of variability in the symbiosis ecology of different coral species. *D. strigosa* individuals could be dominated by four different OTUs, while all *D. cylindrus*, *S. radians*, *M. cavernosa*, and *P. astreoides* individuals were dominated by a single OTU. In fact, this variability, or 'flexibility' in the symbiosis ecology of a host species, can be quantified as multivariate dispersion (a metric of beta diversity) using metabarcoding data, which can be statistically compared among species [@Anderson:2006fs]. The flexibility of symbiosis is important as a potential indicator of the environmental sensitivity of a species [@Putnam:2012wv], and its ability to switch or shuffle its symbionts. This metric revealed that *M. cavernosa*, *P. astreoides*, *S. radians*, and *D. cylindrus* all had the lowest symbiotic flexibility. On the other hand, *D. strigosa*, *P. furcata*, *M. alcicornis*, and *S. siderea* had the most variable symbiont communities, indicating that these species may be more 'flexible' in their associations, and that their symbiont community structure may be more responsive to variability in the natural environment.

@@

  While not providing species-level resolution, ITS2 metabarcoding represents a powerful approach to quantitatively and comprehensively describe *Symbiodinium* metacommunity composition on coral reefs. With appropriate bioinformatic interpretation, these datasets can address critical topics in coral symbiosis ecology, including the influence of the environment on community structure, and changes in dominant or background symbiont abundance over environmental gradients or over time, with important implications for understanding coral responses to environmental change.
